Covariate analysis of CMC RNASeq data reprocessed with Star-Feature counts pipeline (ACC and DLPFC with TMM normlaisation)

### 2016-11-24 code/Rmd/CMC_ACC_And_DLPFC_StarAligned_TMM.Rmd
Added additional metadata (RNAseq technical) to the DLPFC ACC covariates pipeline

# Get merged metadata
METADATA_ID = 'syn2299154'
ALL_USED_IDs[length(ALL_USED_IDs)+1] = METADATA_ID
METADATA = read.csv(synGet(METADATA_ID)@filePath, fill = T, header = F)

@@

## Metadata with specific information for RNASeq processing
metadataVarOfInterest = c("Individual ID", "Institution", "Gender", "Age of Death", "PMI (in hours)", 
                          "Dx", "ACC_RNA_isolation: Exclude?", "ACC_RNA_isolation: Sample RNA ID",
                          "ACC_RNA_isolation: RIN", "ACC_RNA_report: Exclude?", "ACC_RNA_report: Library Batch",
                          "DLPFC_RNA_isolation: Exclude?", "DLPFC_RNA_isolation: Sample RNA ID",
                          "DLPFC_RNA_isolation: RIN", "DLPFC_RNA_report: Exclude?", "DLPFC_RNA_report: Library Batch",
                          'DNA_genotyping_report: Genotyping Sample_ID')

@@

METADATA_ACC = METADATA %>%
  dplyr::select(Individual_ID, Institution, Gender, Age_of_Death, PMI__in_hours_, Dx, ACC_RNA_isolation__Exclude_,
                ACC_RNA_isolation__Sample_RNA_ID, ACC_RNA_isolation__RIN, ACC_RNA_report__Exclude_,
                ACC_RNA_report__Library_Batch, DNA_genotyping_report__Genotyping_Sample_ID) %>%
  dplyr::rename(PMI = PMI__in_hours_, 
                IsolationExclude = ACC_RNA_isolation__Exclude_, 
                SampleID = ACC_RNA_isolation__Sample_RNA_ID, 
                RIN = ACC_RNA_isolation__RIN, 
                ReportExclude = ACC_RNA_report__Exclude_, 
                LibraryBatch = ACC_RNA_report__Library_Batch,
                GenotypingSampleID = DNA_genotyping_report__Genotyping_Sample_ID) %>%
  dplyr::filter(SampleID %in% colnames(COUNT)) %>%
  dplyr::mutate(Tissue = 'ACC')

METADATA_DLPFC = METADATA %>%
  dplyr::select(Individual_ID, Institution, Gender, Age_of_Death, PMI__in_hours_, Dx, DLPFC_RNA_isolation__Exclude_,
                DLPFC_RNA_isolation__Sample_RNA_ID, DLPFC_RNA_isolation__RIN, DLPFC_RNA_report__Exclude_,
                DLPFC_RNA_report__Library_Batch, DNA_genotyping_report__Genotyping_Sample_ID) %>%
  dplyr::rename(PMI = PMI__in_hours_, 
                IsolationExclude = DLPFC_RNA_isolation__Exclude_, 
                SampleID = DLPFC_RNA_isolation__Sample_RNA_ID, 
                RIN = DLPFC_RNA_isolation__RIN, 
                ReportExclude = DLPFC_RNA_report__Exclude_, 
                LibraryBatch = DLPFC_RNA_report__Library_Batch,
                GenotypingSampleID = DNA_genotyping_report__Genotyping_Sample_ID) %>%
  dplyr::filter(SampleID %in% colnames(COUNT)) %>%
  dplyr::mutate(Tissue = 'DLPFC')

METADATA = rbindlist(list(METADATA_ACC, METADATA_DLPFC), use.names = T, fill = T) %>%
  droplevels() %>%
  dplyr::mutate(LibraryBatch = as.factor(LibraryBatch)) %>%
  left_join(ANCESTRY)

@@

### Covariate clustering
Determine relationship between covariates. 
```{r covariates.clustering}
FactorCovariates <- c("Individual_ID", "Institution", "Gender", "LibraryBatch", "Dx.Tissue")
ContCovariates <- c("Age_of_Death", "PMI", "RIN", "EV.1", "EV.2", "EV.3", "EV.4", "EV.5")

# Find inter relation between factor covariates
COVARIATES = METADATA[,c(FactorCovariates,ContCovariates),drop=F]
COVARIATES[,FactorCovariates] <- data.frame(lapply(COVARIATES[,FactorCovariates],function(x){x <- sapply(x,function(y){str_replace_all(as.character(y),'[^[:alnum:]]','_')})}))
rownames(COVARIATES) <- METADATA$SampleID

@@

# Add in RIN^2 values
COVARIATES$RIN2 = COVARIATES$RIN^2
ContCovariates = c(ContCovariates, 'RIN2')
```

@@

```{r dataExplore, fig.width = 6, fig.height = 15}
my.theme <- theme(legend.position = 'top', axis.text.x = element_text(angle = 90, hjust = 1))

# RIN
p = list()
p[[1]] = ggplot(COVARIATES, aes(x = Dx.Tissue, y = RIN)) + geom_boxplot()
p[[1]] = p[[1]] + ggtitle('RIN') + my.theme

# Age of Death
p[[2]] = ggplot(COVARIATES, aes(x = Dx.Tissue, y = Age_of_Death)) + geom_boxplot()
p[[2]] = p[[2]] + ggtitle('AgeOfDeath') + my.theme

# PMI
p[[3]] = ggplot(COVARIATES, aes(x = Dx.Tissue, y = PMI)) + geom_boxplot()
p[[3]] = p[[3]] + ggtitle('PMI (in hours)') + my.theme

multiplot(plotlist = p, cols = 1)
```
### CPM Normalisation
Preprocess counts matrix and metadata. Determine design matrix for normalisation and differential expression analysis.

@@

```{r get.gene.symbols, cache=TRUE, eval=T}
backgroundGenes = data.frame(ensembl_gene_id = rownames(COUNT))

# Query biomart
Ensemble2HGNC <- getBM(attributes = c("ensembl_gene_id", "hgnc_symbol"),
                       filters = "ensembl_gene_id", values = backgroundGenes$ensembl_gene_id,
                       mart = mart)

# Add gene names
RESIDUAL.GENE_EXPRESSION <- RESIDUAL.GENE_EXPRESSION %>%
  rownameToFirstColumn('ensembl_gene_id') %>%
  left_join(Ensemble2HGNC) %>%
  dplyr::select(ensembl_gene_id, hgnc_symbol, one_of(rownames(COVARIATES)))

>>>>

# Get merged metadata
METADATA_ID = 'syn7574833';#'syn2299154'
ALL_USED_IDs[length(ALL_USED_IDs)+1] = METADATA_ID
METADATA = read.csv(synGet(METADATA_ID)@filePath, fill = T, header = F)

@@

## Metadata with specific information for RNASeq processing
metadataVarOfInterest = c("Individual ID", "Institution", "Gender", "Age of Death", "PMI (in hours)", "Dx",

                          "ACC_RNA_isolation: Exclude?", "ACC_RNA_isolation: Sample RNA ID",
                          "ACC_RNA_isolation: RIN", "ACC_RNA_report: Exclude?", "ACC_RNA_report: Library Batch",
                          "ACC_RNA_report: Ribozero Batch", "ACC_RNA_report: Flowcell Batch", 
                          "ACC_RNA_report: Mapped Reads", "ACC_RNA_report: Intragenic Rate",
                          "ACC_RNA_report: Intronic Rate", "ACC_RNA_report: Intergenic Rate",
                          "ACC_RNA_report: Genes Detected", "ACC_RNA_report: Transcripts Detected",
                          "ACC_RNA_report: Expression Profiling Efficiency", "ACC_RNA_report: rRNA Rate",
                          "ACC_RNA_report: Total Reads", "ACC_RNA_report: Percent Aligned",

                          "DLPFC_RNA_isolation: Exclude?", "DLPFC_RNA_isolation: Sample RNA ID",
                          "DLPFC_RNA_isolation: RIN", "DLPFC_RNA_report: Exclude?", "DLPFC_RNA_report: Library Batch",
                          "DLPFC_RNA_report: Ribozero Batch", "DLPFC_RNA_report: Flowcell Batch",
                          "DLPFC_RNA_report: Mapped Reads", "DLPFC_RNA_report: Intragenic Rate",
                          "DLPFC_RNA_report: Intronic Rate", "DLPFC_RNA_report: Intergenic Rate", 
                          "DLPFC_RNA_report: Genes Detected", "DLPFC_RNA_report: Transcripts Detected", 
                          "DLPFC_RNA_report: Expression Profiling Efficiency", "DLPFC_RNA_report: rRNA Rate",
                          "DLPFC_RNA_report: Total Reads", "DLPFC_RNA_report: Percent Aligned",

                          "DNA_genotyping_report: Genotyping Sample_ID")

@@

METADATA_ACC = METADATA %>%
  dplyr::select(Individual_ID, Institution, Gender, Age_of_Death, PMI__in_hours_, Dx, 
                ACC_RNA_isolation__Exclude_, ACC_RNA_isolation__Sample_RNA_ID, ACC_RNA_isolation__RIN,
                ACC_RNA_report__Exclude_, ACC_RNA_report__Ribozero_Batch, ACC_RNA_report__Library_Batch, 
                ACC_RNA_report__Flowcell_Batch, ACC_RNA_report__Mapped_Reads, ACC_RNA_report__Intragenic_Rate,
                ACC_RNA_report__Intronic_Rate, ACC_RNA_report__Intergenic_Rate, ACC_RNA_report__Genes_Detected,
                ACC_RNA_report__Transcripts_Detected, ACC_RNA_report__Expression_Profiling_Efficiency, 
                ACC_RNA_report__rRNA_Rate, ACC_RNA_report__Total_Reads, ACC_RNA_report__Percent_Aligned,
                DNA_genotyping_report__Genotyping_Sample_ID) %>%
  dplyr::rename(PMI = PMI__in_hours_, 
                IsolationExclude = ACC_RNA_isolation__Exclude_, 
                SampleID = ACC_RNA_isolation__Sample_RNA_ID, 
                RIN = ACC_RNA_isolation__RIN, 
                ReportExclude = ACC_RNA_report__Exclude_, 
                RibozeroBatch = ACC_RNA_report__Ribozero_Batch,
                LibraryBatch = ACC_RNA_report__Library_Batch,
                FlowcellBatch = ACC_RNA_report__Flowcell_Batch, 
                MappedReads = ACC_RNA_report__Mapped_Reads, 
                IntragenicRate = ACC_RNA_report__Intragenic_Rate,
                IntronicRate = ACC_RNA_report__Intronic_Rate, 
                IntergenicRate = ACC_RNA_report__Intergenic_Rate, 
                GenesDetected = ACC_RNA_report__Genes_Detected,
                TranscriptsDetected = ACC_RNA_report__Transcripts_Detected, 
                ExpProfEfficiency = ACC_RNA_report__Expression_Profiling_Efficiency, 
                rRNARate = ACC_RNA_report__rRNA_Rate, 
                TotalReads = ACC_RNA_report__Total_Reads, 
                PercentAligned = ACC_RNA_report__Percent_Aligned,
                GenotypingSampleID = DNA_genotyping_report__Genotyping_Sample_ID) %>%
  dplyr::filter(SampleID %in% colnames(COUNT)) %>%
  dplyr::mutate(Tissue = 'ACC')

METADATA_DLPFC = METADATA %>%
  dplyr::select(Individual_ID, Institution, Gender, Age_of_Death, PMI__in_hours_, Dx, 
                DLPFC_RNA_isolation__Exclude_, DLPFC_RNA_isolation__Sample_RNA_ID, DLPFC_RNA_isolation__RIN,
                DLPFC_RNA_report__Exclude_, DLPFC_RNA_report__Ribozero_Batch, DLPFC_RNA_report__Library_Batch, 
                DLPFC_RNA_report__Flowcell_Batch, DLPFC_RNA_report__Mapped_Reads, DLPFC_RNA_report__Intragenic_Rate,
                DLPFC_RNA_report__Intronic_Rate, DLPFC_RNA_report__Intergenic_Rate, DLPFC_RNA_report__Genes_Detected,
                DLPFC_RNA_report__Transcripts_Detected, DLPFC_RNA_report__Expression_Profiling_Efficiency, 
                DLPFC_RNA_report__rRNA_Rate, DLPFC_RNA_report__Total_Reads, DLPFC_RNA_report__Percent_Aligned,
                DNA_genotyping_report__Genotyping_Sample_ID) %>%
  dplyr::rename(PMI = PMI__in_hours_, 
                IsolationExclude = DLPFC_RNA_isolation__Exclude_, 
                SampleID = DLPFC_RNA_isolation__Sample_RNA_ID, 
                RIN = DLPFC_RNA_isolation__RIN, 
                ReportExclude = DLPFC_RNA_report__Exclude_, 
                RibozeroBatch = DLPFC_RNA_report__Ribozero_Batch,
                LibraryBatch = DLPFC_RNA_report__Library_Batch,
                FlowcellBatch = DLPFC_RNA_report__Flowcell_Batch, 
                MappedReads = DLPFC_RNA_report__Mapped_Reads, 
                IntragenicRate = DLPFC_RNA_report__Intragenic_Rate,
                IntronicRate = DLPFC_RNA_report__Intronic_Rate, 
                IntergenicRate = DLPFC_RNA_report__Intergenic_Rate, 
                GenesDetected = DLPFC_RNA_report__Genes_Detected,
                TranscriptsDetected = DLPFC_RNA_report__Transcripts_Detected, 
                ExpProfEfficiency = DLPFC_RNA_report__Expression_Profiling_Efficiency, 
                rRNARate = DLPFC_RNA_report__rRNA_Rate, 
                TotalReads = DLPFC_RNA_report__Total_Reads, 
                PercentAligned = DLPFC_RNA_report__Percent_Aligned,
                GenotypingSampleID = DNA_genotyping_report__Genotyping_Sample_ID) %>%
  dplyr::filter(SampleID %in% colnames(COUNT)) %>%
  dplyr::mutate(Tissue = 'DLPFC')

METADATA = rbindlist(list(METADATA_ACC, METADATA_DLPFC), use.names = T, fill = T) %>%
  droplevels() %>%
  dplyr::mutate(LibraryBatch = as.factor(LibraryBatch)) %>%
  left_join(ANCESTRY) %>%
  unique

@@

### Covariate clustering
Determine relationship between covariates. 
```{r covariates.clustering}
FactorCovariates <- c('Individual_ID', "Institution", "Gender", "RibozeroBatch", "LibraryBatch", "FlowcellBatch", "Dx.Tissue")
ContCovariates <- c("Age_of_Death", "PMI", "RIN", "MappedReads", "IntragenicRate", "IntronicRate", "IntergenicRate",
                    "GenesDetected", "TranscriptsDetected", "ExpProfEfficiency", "rRNARate", "TotalReads", 
                    "PercentAligned", "EV.1", "EV.2", "EV.3", "EV.4", "EV.5")

# Find inter relation between factor covariates
COVARIATES = METADATA[,c(FactorCovariates,ContCovariates),drop=F]
COVARIATES[,FactorCovariates] <- data.frame(lapply(COVARIATES[,FactorCovariates],function(x){
  x <- sapply(x,function(y){str_replace_all(as.character(y),'[^[:alnum:]]','_')})}))
rownames(COVARIATES) <- METADATA$SampleID

@@

# Add in RIN^2 values
COVARIATES$RIN2 = COVARIATES$RIN^2
ContCovariates = c(ContCovariates, 'RIN2')

## Get GC content from biomart
backgroundGenes = data.frame(ensembl_gene_id = rownames(COUNT))

# Query biomart
Ensemble2HGNC <- getBM(attributes = c("ensembl_gene_id", "hgnc_symbol", "percentage_gc_content"),
                       filters = "ensembl_gene_id", values = backgroundGenes$ensembl_gene_id,
                       mart = mart)
GC.CONT = Ensemble2HGNC %>%
  dplyr::select(ensembl_gene_id, percentage_gc_content) %>%
  unique
rownames(GC.CONT) = GC.CONT$ensembl_gene_id
```

@@

```{r dataExplore, fig.width = 5, fig.height = 5}
my.theme <- theme(legend.position = 'top', axis.text.x = element_text(angle = 90, hjust = 1))

# RIN
p = list()
p[[1]] = ggplot(COVARIATES, aes(x = Dx.Tissue, y = RIN)) + geom_boxplot()
p[[1]] = p[[1]] + ggtitle('RIN') + my.theme
p[[1]]

# Age of Death
p[[2]] = ggplot(COVARIATES, aes(x = Dx.Tissue, y = Age_of_Death)) + geom_boxplot()
p[[2]] = p[[2]] + ggtitle('AgeOfDeath') + my.theme
p[[2]]

# PMI
p[[3]] = ggplot(COVARIATES, aes(x = Dx.Tissue, y = PMI)) + geom_boxplot()
p[[3]] = p[[3]] + ggtitle('PMI (in hours)') + my.theme
p[[3]]

# Intronic Rate
p[[4]] = ggplot(COVARIATES, aes(x = Dx.Tissue, y = IntronicRate)) + geom_boxplot()
p[[4]] = p[[4]] + ggtitle('Intronic Rate') + my.theme
p[[4]]

# IntergenicRate
p[[5]] = ggplot(COVARIATES, aes(x = Dx.Tissue, y = IntergenicRate)) + geom_boxplot()
p[[5]] = p[[5]] + ggtitle('Intergenic Rate') + my.theme
p[[5]]

# Transcripts Detected
p[[6]] = ggplot(COVARIATES, aes(x = Dx.Tissue, y = TranscriptsDetected)) + geom_boxplot()
p[[6]] = p[[6]] + ggtitle('Transcripts Detected') + my.theme
p[[6]]

# Mapped Reads
p[[7]] = ggplot(COVARIATES, aes(x = Dx.Tissue, y = MappedReads)) + geom_boxplot()
p[[7]] = p[[7]] + ggtitle('Mapped Reads') + my.theme
p[[7]]

# PercentAligned
p[[8]] = ggplot(COVARIATES, aes(x = Dx.Tissue, y = PercentAligned)) + geom_boxplot()
p[[8]] = p[[8]] + ggtitle('Percent Aligned') + my.theme
p[[8]]

# rRNARate
p[[9]] = ggplot(COVARIATES, aes(x = Dx.Tissue, y = rRNARate)) + geom_boxplot()
p[[9]] = p[[9]] + ggtitle('rRNARate') + my.theme
p[[9]]

# Institution
vcd::mosaic(~ Institution + Dx.Tissue, data = COVARIATES)

# Gender
vcd::mosaic(~ Gender + Dx.Tissue, data = COVARIATES)
```
### CPM Normalisation
Preprocess counts matrix and metadata. Determine design matrix for normalisation and differential expression analysis.

@@

```{r get.gene.symbols, cache=TRUE, eval=T}
# Add gene names
RESIDUAL.GENE_EXPRESSION <- RESIDUAL.GENE_EXPRESSION %>%
  rownameToFirstColumn('ensembl_gene_id') %>%
  left_join(Ensemble2HGNC) %>%
  dplyr::select(ensembl_gene_id, hgnc_symbol, one_of(rownames(COVARIATES)))
